Explaining probabilistic prediction on the simplex with Shapley compositions

### 2023-12-19 paper/main.tex
edit paper

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

@@

An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built. The projection of a composition (a discrete probability distribution in our case) into this basis defined an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$. This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry. Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition, multiplication by a scalar and inner product.

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed in Section \ref{sec:moreclasses}.

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

\subsubsection{More classes}
\label{sec:moreclasses}
When there are more classes, all the dimension of the ILR space cannot be visualized at once but $2$ or $3$-dimensional subspaces can still be visualized. However, in order to choose which ILR components to visualize, one needs to understand what they refer too. The next section provides some intuition about that in the context of the explanation of a classifier prediction with more than 4 classes.

\subsection{Groups of part, balances and sequential binary partition}
\ref{sec:balances}
..TODO..

@@

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction. It gives the overall contribution of the feature on the prediction, regardless of its direction. The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.

>>>>

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

@@

An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built. The projection of a composition (like a discrete probability distribution) into this basis defines an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$. This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry. Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition, multiplication by a scalar and inner product.

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed in Section \ref{sec:balances}.

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

\subsection{More classes: groups of parts and balances}
\label{sec:balances}

When there are more classes, all the dimension of the ILR space cannot be visualized at once. However, $2$ or $3$-dimensional subspaces can still be visualized. In order to choose which ILR components to visualize, one needs to understand what they refer to. In this section, we briefly discuss the interpretation of the ILR components in the context of the explanation of a classifier prediction with more than 4 classes.

The projection of a composition into an element of the Aitchison orthonormal basis, i.e. a component of the ILR space. This can be interpreted as a \emph{balance}, i.e. a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}. One giving the central values of the parts in one group of parts and one for another group of parts. Therefore, in our context, a balances is comparing the weight of two group of classes. This can be illustrated by a sequential binary partition of bifurcation tree. Figure \ref{fig:trees} gives two examples.
\begin{figure}
  \centering
  \subfigure[]{\label{fig:bifurc1}
    \input{bifurctree.tex}
    }
  \subfigure[]{\label{fig:bifurc2}
    \input{bifurctree.tex}
  }
  \caption{Two examples of bifurcation tree. \ref{fig:bifurc1} is the one corresponding to the basis obtained by with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}...}
  \label{fig:trees}
\end{figure}

Geometric orthogonality (not statistical uncorrelation), balances provide nonredundant information about the composition.\\
Balancing elements\\
\cite{egozcue2005groups}

@@

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction. It gives the overall contribution of the feature on the prediction, regardless of its direction. The angle between two Shapley compositions can informs about their orthogonality. If two Shapley composition are orthogonal\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}, this suggest the features are nonredundant. A negative angle would suggest that the features have an opposite influence on the prediction. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.
